\bf Baseline Drift Estimation for Time Series Data Using Quantile Trend Filtering

### 2019-03-06 Manuscript/detrendify_JCGS.tex

@@

%% Vectors
\newcommand{\V}[1]{{\bm{\mathbf{\MakeLowercase{#1}}}}} % vector
\newcommand{\VE}[2]{\MakeLowercase{#1}_{#2}} % vector element
\newcommand{\Vbar}[1]{{\bm{\M{A}r \mathbf{\MakeLowercase{#1}}}}} % vector
\newcommand{\Vhat}[1]{{\bm{\hat \mathbf{\MakeLowercase{#1}}}}} % vector
\newcommand{\Vtilde}[1]{{\bm{\tilde \mathbf{\MakeLowercase{#1}}}}} % vector
\newcommand{\Vn}[2]{\V{#1}^{(#2)}} % n-th vector
\newcommand{\VnE}[3]{{#1}^{(#2)}_{#3}} % n-th vector
\newcommand{\VtildeE}[2]{\tilde{\MakeLowercase{#1}}_{#2}} % vector element

%% Matrices
\newcommand{\M}[1]{{\bm{\mathbf{\MakeUppercase{#1}}}}} % matrix
\newcommand{\ME}[2]{\MakeLowercase{#1}_{#2}} % matrix element
\newcommand{\MC}[2]{\V{#1}_{#2}}
\newcommand{\Mr}[2]{\V{#1}_{#2}} % matrix row
\newcommand{\Mhat}[1]{{\bm{\hat \mathbf{\MakeUppercase{#1}}}}} % matrix
\newcommand{\Mtilde}[1]{{\bm{\tilde \mathbf{\MakeUppercase{#1}}}}} % matrix
\newcommand{\MhatC}[2]{\Vhat{#1}_{#2}} % matrix column
\newcommand{\Mbar}[1]{{\bm{\M{A}r \mathbf{\MakeUppercase{#1}}}}} % matrix
\newcommand{\MbarC}[2]{\Vbar{#1}_{#2}} % matrix column
\newcommand{\Mn}[2]{\M{#1}^{(#2)}} % n-th matrix
\newcommand{\Mbarn}[2]{\Mbar{#1}^{(#2)}} % n-th matrix
\newcommand{\Mtilden}[2]{\Mtilde{#1}^{(#2)}} % n-th matrix
\newcommand{\MnTra}[2]{\M{#1}^{(#2){{\sf T}}}} % n-th matrix transpose
\newcommand{\MnE}[3]{\MakeLowercase{#1}^{(#2)}_{#3}} % n-th matrix element
\newcommand{\MnC}[3]{\V{#1}^{(#2)}_{#3}} % n-th matrix column
\newcommand{\MbarnC}[3]{\Vbar{#1}^{(#2)}_{#3}} % n-th matrix column
\newcommand{\MnCTra}[3]{\V{#1}^{(#2){{\sf T}}}_{#3}} % n-th matrix column transpose
\newcommand{\MCTra}[2]{\V{#1}^{{\sf T}}_{#2}} % n-th matrix column transpose

@@

	In a wide range of applications that spans chemistry \citep{Ning2014}, macroeconomics \citep{yamada2017estimating}, environmental science \citep{brantley2014mobile}, and medical sciences \citep{pettersson2013algorithm, marandi2015qualitative}, scalar functions of time $y(t)$ are observed and assumed to be a superposition of an underlying slowly varying trend $\theta(t)$, other more rapidly varying components $s(t)$, and noise. In practice, $y(t)$ is observed at discrete time points $t_1, \ldots, t_n$, and we model the vector of signal samples $\VE{y}{i} = y(t_i)$ as	
\begin{eqnarray*}
\V{y} & = & \V{\theta} + \V{s} + \V{\varepsilon},
\end{eqnarray*}
where $\VE{\theta}{i} = \theta(t_i)$, $\VE{s}{i} = s(t_i)$, and $\V{\varepsilon} \in \Real^n$ is a vector of uncorrelated noise. For notational simplicity, for the rest of the paper, we assume that the time points take on the values $t_i = i$, but it is straightforward to generalize to an arbitrary grid of time points.

In some applications, the slowly varying component $\V{\theta}$ is the signal of interest and the transient component $\V{s}$ is a vector of nuisance parameters. In other applications, the roles of $\V{\theta}$ and $\V{s}$ are reversed. The latter applies in the air quality applications motivating this work, as $\V{s}$ represents the signal of interest and $\V{\theta}$ represents a baseline drift that obscures the identification of the important transient events encoded in $\V{s}$. In either case, the estimation task is to demix the noisy observed vector $\V{y}$ into a slowly varying component $\V{\theta}$ and a transient component $\V{s}$.

To tackle this demixing problem, we introduce a scalable baseline estimation framework by building on \textit{$\ell_1$-trend filtering}, a relatively new nonparametric estimation framework. \cite{Kim2009} originally proposed \textit{$\ell_1$-trend filtering} to estimate trends with piecewise polynomial functions, assuming that the observed time series $\V{y}$ consists of a trend $\V{\theta}$ plus uncorrelated noise $\V{\varepsilon}$, namely $\V{y} = \V{\theta} + \V{\varepsilon}$. The estimated trend is the solution to the following convex optimization problem
\begin{eqnarray*}
\underset{\V{\theta}}{\text{minimize}}\; \frac{1}{2} \lVert \V{y} - \V{\theta} \rVert_2^2 + \lambda \lVert \Mn{D}{k+1}\V{\theta} \rVert_1,
\end{eqnarray*}
	where $\lambda$ is a nonnegative regularization parameter, and the matrix $\Mn{D}{k+1} \in \Real^{(n - k -1) \times n}$ is the discrete difference operator of order $k+1$. To understand the purpose of penalizing $\Mn{D}{k+1}$ consider the difference operator when $k = 0$.
	\begin{eqnarray*}
	\Mn{D}{1} & = & \begin{pmatrix}

	-1 & 1 & 0 & \cdots & 0 & 0 \\
	0 & -1 & 1 & \cdots & 0 & 0 \\
	\vdots & & & & & \\
	0 & 0 & 0 & \cdots & -1 & 1 \\

	\end{pmatrix}.
	\end{eqnarray*}
	Thus, $\lVert \Mn{D}{1}\V{\theta} \rVert_1 = \sum_{i=1}^{n-1} \lvert \theta_i - \theta_{i+1} \rvert$ which is just total variation denoising in one dimension. The penalty term incentivizes solutions which are piecewise constant. For $k \geq 1$, the difference operator $\Mn{D}{k+1} \in \Real^{(n-k-1) \times n}$ is defined recursively as follows
	\begin{eqnarray*}
	\Mn{D}{k+1} & = & \Mn{D}{1}\Mn{D}{k}.
	\end{eqnarray*}
	Penalizing the 1-norm of the vector $\Mn{D}{k+1}\V{\theta}$ produces estimates of $\V{\theta}$ that are piecewise polynomials of order $k$.

@@

	\cite{Tib2014} proved that with a judicious choice of $\lambda$ the trend filtering estimate converges to the true underlying function at the minimax rate for functions whose $k$th derivative is of bounded variation and showed that trend filtering is both fast and locally adaptive when the time series consists of only the trend and random noise, which is illustrated in \Fig{tpn}. As noted earlier though in some applications, such as the air quality monitoring problem considered in this paper, the data contain a rapidly varying signal in addition to the slowly varying trend and noise. Standard trend filtering struggles to distinguish between the slowly varying trend and the rapidly-varying signal, illustrated in \Fig{tpspn}.

	To account for the presence of a transient component in the observed time series $\V{y}$, we propose quantile trend filtering \Fig{tpspn}.% as a more robust method for detrending in these situations \Fig{tpspn}. 
	To estimate the trend in the $\tau$\textsuperscript{th} quantile, we solve the convex optimization problem
	\begin{eqnarray}
	\label{eq:quantile_trend}
	\underset{\V{\theta}}{\text{minimize}}\; \rho_\tau(\V{y} - \V{\theta}) + \lambda \lVert \Mn{D}{k+1} \V{\theta} \rVert_1,
	\end{eqnarray}
	where $\rho_\tau(\V{r})$ is the check function
	\begin{eqnarray}
	\label{eq:check}
	\rho_{\tau}(\V{r}) & = & \sum_{i=1}^n \VE{r}{i}(1-\One(\VE{r}{i}<0)),
	\end{eqnarray}
	and $\One(A)$ is 1 if its input $A$ is true and 0 otherwise. \cite{Kim2009} proposed using the check function as a possible extension of $\ell_1$-trend filtering but did not investigate it further. Here, we develop the basic $\ell_1$-quantile-trend-filtering framework and extend it to model multiple quantiles simultaneously with non-crossing constraints to ensure valid trend estimates. To reduce computation time and extend the method to larger series, we develop and a parallelizable ADMM algorithm. Finally, we propose a modified criterion for choosing the smoothing parameter.

	\begin{figure}
		\centering
		\subfloat[Slow Trend + Noise]{\label{fig:tpn}
		\includegraphics[width = 0.45\linewidth]{Figures/trend_filter_eg1.png}}
		\subfloat[Slow Trend + Spikes + Noise]{\label{fig:tpspn}
		\includegraphics[width = 0.45\linewidth]{Figures/trend_filter_eg2.png}}
		\caption{Examples of trend filtering solutions (red) and 15\textsuperscript{th} quantile trending filtering solution (blue). Standard trend filtering performs well in the no-signal case (right) but struggles to distinguish between the slowly varying trend and the rapidly-varying signal (left). The quantile trend is not affected by the signal and provides an estimate of the baseline.}
	\end{figure}

	This work was motivated by the analysis of signals measured by low cost and portable air quality sensors, whose use has increased dramatically in the last decade. These sensors can provide an un-calibrated measure of a variety of pollutants in near real time, but deriving meaningful information from sensor data remains a challenge \citep{snyder2013changing}. The ``SPod" is a low-cost sensor currently being investigated by researchers at the U.S. Environmental Protection Agency to detect volatile organic compound (VOC) emissions from industrial facilities \citep{thoma2016south}. To reduce cost and power consumption of the SPod, the temperature and relative humidity of the air presented to the photoionization detectors (PIDs) is not controlled and as a result the output signal exhibits a slowly varying baseline drift on the order of minutes to hours. Figure \ref{fig:raw_spod} provides an example of measurements from three SPod sensors co-located at the border of an industrial facility. All of the sensors respond to the pollutant signal, the three sharp transient spikes at \Halley{Eric}{can you put the times at which the spikes occur here?}, but the amount of baseline drift varies from one sensor to another obscuring the detection of the peaks that herald the presence of pollutants. We show later that by estimating the baseline drift in each sensor using $\ell_1$-quantile-trend filtering and removing it from the observed signals, peaks can be reliably detected from concordant residual signals from a collection of SPods using a simple data-driven thresholding strategy.

@@

	\begin{equation*}
	\label{eq:smoothingspline}
	\sum_{t=1}^n\rho_{\tau}(y(t) - \theta(t)) + \lambda\int \theta''(t)^2 dt, 
 	\end{equation*}	
	where $\theta(t)$ is a smooth function of time and $\lambda$ is a tuning parameter that controls the degree of smoothing. \cite{Oh2011} proposed an algorithm for solving the quantile smoothing spline problem by approximating the check function with a similar differentiable function. \cite{Racine2017} take a different approach and constrain the response to follow a smooth location scale model of the form $y(t) = a(t) + b(t)\epsilon_i$ and estimate the $\tau$\textsuperscript{th} conditional quantile using a kernel estimator and local linear approach.

%% ----------------------------------------------------------------------
%% Methods
%% ----------------------------------------------------------------------
\section{Methods}
\label{sec:methods}
%% ----------------------------------------------------------------------
%% Quantile Trend Filtering
%% ----------------------------------------------------------------------
\subsection{Quantile Trend Filtering}

	We combine the ideas of quantile regression and trend filtering. For a single quantile level $\tau$, the quantile trend filtering problem is given in \Eqn{quantile_trend}. As with classic quantile regression, the quantile trend filtering problem is a linear program which can be solved by a number of methods. In many cases, including ours, we are interested in estimating multiple quantiles simultaneously. We also want to ensure that our quantile estimates are valid by enforcing the constraint that if $\tau_2 > \tau_1$ then $Q(\tau_2) \ge Q(\tau_1)$ where $Q$ is the quantile function of $\V{y}$. Given quantiles $\tau_1 < \tau_2 < \ldots < \tau_J$, the optimization problem becomes 
	\begin{eqnarray*}
	\label{eq:noncross_trend}
%	&&\underset{\V{\theta}_1, \ldots, \V{\theta}_J}{\text{minimize}}\; \sum_{j=1}^J \left [\rho_{\tau_j}(\V{y} - \V{\theta}_{j}) + 
%	\lambda_j \lVert \Mn{D}{k+1} \V{\theta}_j \rVert_1 \right ] \\	
	&&\underset{\M{\Theta} \in \mathcal{C}}{\text{minimize}}\; \sum_{j=1}^J \left [\rho_{\tau_j}(\V{y} - \V{\theta}_{j}) + 
	\lambda_j \lVert \Mn{D}{k+1} \V{\theta}_j \rVert_1 \right ] \\
%	&&\text{subject to: }\; \V{\theta}_1 \leq \V{\theta}_2 \leq \ldots \leq \V{\theta}_J	
%	\theta_{1}(t) \le \theta_{2}(t) \le \ldots \le \theta_{J}(t) \text{ for all } t, \nonumber
	\end{eqnarray*}
	where $\M{\Theta} = \begin{pmatrix} \V{\theta}_1 & \V{\theta}_2 & \cdots & \V{\theta}_J \end{pmatrix} \in \Real^{n \times J}$ is a matrix whose $j$th column corresponds to the $j$th quantile signal $\V{\theta}_j$ and the set $\mathcal{C} = \{\M{\Theta} \in \Real^{n \times J}: \ME{\theta}{ij} \leq \ME{\theta}{ij'} \;\text{for $j \leq j'$}\}$ encodes the non-crossing quantile constraints. The additional constraints are linear in the parameters, so the non-crossing quantile trends can still be estimated by a number of available solvers. In the rest of this paper, we report numerical results using the commercial solver Gurobi \citep{gurobi} and its R package implementation. However, we could easily substitute a free solver such as the Rglpk package by \cite{rglpk}.

@@

	The number of parameters to be estimated in this problem is equal to the number of observations multiplied by the number of quantiles of interest. As the size of the data and the number of quantiles grows, all solvers will eventually break \Halley{Eric}{Can we be more specific about what we mean when we say ``break?"}. To our knowledge, no one has addressed the problem of finding smooth quantile trends of series that are too large to be processed simultaneously. We propose an ADMM algorithm for solving large problems in a piecewise fashion. The ADMM algorithm \citep{gabay1975dual, glowinski1975approximation} is fully described by \cite{boyd2011distributed}.

	Let $\mathcal{I}$ denote a subset of the indices $\{1, \ldots, n\}$ and $\V{y}_\mathcal{I} \in \Real^{\lvert I \rvert}$ denote the subvector obtained by stacking on top of each other the elements $\VE{y}{i}$ for $i \in \mathcal{I}$ in their canonical ordering.	
	We apply the consensus ADMM algorithm to the quantile regression trend filtering problem given in \Eqn{quantile_trend}, by dividing our observed series $\VE{y}{i}$ with $i = 1, \ldots, n$ into overlapping $M$ windows of observations $\Vn{y}{m} = \V{y}_{\mathcal{I}_m}$ where $\mathcal{I}_m$ is the set of integers from $l_m$ to $u_m$ inclusive, namely $\mathcal{I}_m = \{l_m, l_m + 1, \ldots, u_m -1, u_m\}$ where $1 = l_{1} < l_{2} < u_{1} < l_{3} < u_{2} < l_{4} < u_{3} < \cdots < u_{M} = n$. \Fig{windows} shows an example of 1200 observations being mapped into three equally sized overlapping windows of observations.

>>>>

\newtheorem{proposition}{Proposition}[section]

@@

%% Vectors
\newcommand{\V}[1]{{\bm{\mathbf{\MakeLowercase{#1}}}}} % vector
\newcommand{\VE}[2]{\MakeLowercase{#1}_{#2}} % vector element
\newcommand{\Vbar}[1]{{\bm{\bar \mathbf{\MakeLowercase{#1}}}}} % vector
\newcommand{\Vhat}[1]{{\bm{\hat \mathbf{\MakeLowercase{#1}}}}} % vector
\newcommand{\Vtilde}[1]{{\bm{\tilde \mathbf{\MakeLowercase{#1}}}}} % vector
\newcommand{\Vn}[2]{\V{#1}^{(#2)}} % n-th vector
\newcommand{\VnE}[3]{{#1}^{(#2)}_{#3}} % n-th vector
\newcommand{\VtildeE}[2]{\tilde{\MakeLowercase{#1}}_{#2}} % vector element

%% Matrices
\newcommand{\M}[1]{{\bm{\mathbf{\MakeUppercase{#1}}}}} % matrix
\newcommand{\ME}[2]{\MakeLowercase{#1}_{#2}} % matrix element
\newcommand{\MC}[2]{\V{#1}_{#2}}
\newcommand{\Mr}[2]{\V{#1}_{#2}} % matrix row
\newcommand{\Mhat}[1]{{\bm{\hat \mathbf{\MakeUppercase{#1}}}}} % matrix
\newcommand{\Mtilde}[1]{{\bm{\tilde \mathbf{\MakeUppercase{#1}}}}} % matrix
\newcommand{\MhatC}[2]{\Vhat{#1}_{#2}} % matrix column
\newcommand{\Mbar}[1]{{\bm{bar \mathbf{\MakeUppercase{#1}}}}} % matrix
\newcommand{\MbarC}[2]{\Vbar{#1}_{#2}} % matrix column
\newcommand{\Mn}[2]{\M{#1}^{(#2)}} % n-th matrix
\newcommand{\Mbarn}[2]{\Mbar{#1}^{(#2)}} % n-th matrix
\newcommand{\Mtilden}[2]{\Mtilde{#1}^{(#2)}} % n-th matrix
\newcommand{\MnTra}[2]{\M{#1}^{(#2){{\sf T}}}} % n-th matrix transpose
\newcommand{\MnE}[3]{\MakeLowercase{#1}^{(#2)}_{#3}} % n-th matrix element
\newcommand{\MnC}[3]{\V{#1}^{(#2)}_{#3}} % n-th matrix column
\newcommand{\MbarnC}[3]{\Vbar{#1}^{(#2)}_{#3}} % n-th matrix column
\newcommand{\MnCTra}[3]{\V{#1}^{(#2){{\sf T}}}_{#3}} % n-th matrix column transpose
\newcommand{\MCTra}[2]{\V{#1}^{{\sf T}}_{#2}} % n-th matrix column transpose

@@

	In a wide range of applications that spans chemistry \citep{Ning2014}, macroeconomics \citep{yamada2017estimating}, environmental science \citep{brantley2014mobile}, and medical sciences \citep{pettersson2013algorithm, marandi2015qualitative}, scalar functions of time $y(t)$ are observed and assumed to be a superposition of an underlying slowly varying baseline trend $\theta(t)$, other more rapidly varying components $s(t)$, and noise. In practice, $y(t)$ is observed at discrete time points $t_1, \ldots, t_n$, and we model the vector of signal samples $\VE{y}{i} = y(t_i)$ as	
\begin{eqnarray*}
\V{y} & = & \V{\theta} + \V{s} + \V{\varepsilon},
\end{eqnarray*}
where $\VE{\theta}{i} = \theta(t_i)$, $\VE{s}{i} = s(t_i)$, and $\V{\varepsilon} \in \Real^n$ is a vector of uncorrelated noise. For notational simplicity, for the rest of the paper, we assume that the time points take on the values $t_i = i$, but it is straightforward to generalize to an arbitrary grid of time points.

In some applications, the slowly varying component $\V{\theta}$ is the signal of interest and the transient component $\V{s}$ is a vector of nuisance parameters. In other applications, the roles of $\V{\theta}$ and $\V{s}$ are reversed. The latter applies in the air quality applications motivating this work, as $\V{s}$ represents the signal of interest and $\V{\theta}$ represents a baseline drift that obscures the identification of the important transient events encoded in $\V{s}$. We describe this motivating problem further to illustrate the critical importance of being able to separate $\V{s}$ from $\V{\theta}$.

%	This work was motivated by the analysis of signals measured by low cost and portable air quality sensors, whose use has increased dramatically in the last decade. 
In the last decade, low cost and portable air quality sensors have enjoyed dramatically increased usage. These sensors can provide an un-calibrated measure of a variety of pollutants in near real time, but deriving meaningful information from sensor data remains a challenge \citep{snyder2013changing}. The ``SPod" is a low-cost sensor currently being investigated by researchers at the U.S. Environmental Protection Agency to detect volatile organic compound (VOC) emissions from industrial facilities \citep{thoma2016south}. To reduce cost and power consumption of the SPod, the temperature and relative humidity of the air presented to the photoionization detectors (PIDs) is not controlled and as a result the output signal exhibits a slowly varying baseline drift on the order of minutes to hours. Figure \ref{fig:raw_spod} provides an example of measurements from three SPod sensors co-located at the border of an industrial facility. All of the sensors respond to the pollutant signal, the three sharp transient spikes at \Halley{Eric}{can you put the times at which the spikes occur here?}, but the baseline drift varies from one sensor to another obscuring the detection of the peaks that alert the intrusion of pollutants. We show later that by estimating the baseline drift in each sensor %using $\ell_1$-quantile-trend filtering 
 and  removing it from the observed signals, peaks can be reliably detected from concordant residual signals from a collection of SPods using a simple data-driven thresholding strategy. Thus, accurately demix a noisy observed vector $\V{y}$ into a slowly varying component $\V{\theta}$ and a transient component $\V{s}$ can lead to greatly improved and simplified downstream analysis. This is true for not just air quality applications, but universally across the many domains listed earlier.

To tackle demixing problems, we introduce a scalable baseline estimation framework by building on \textit{$\ell_1$-trend filtering}, a relatively new nonparametric estimation framework. Our contributions are three-fold. 
\begin{itemize}
\item \cite{Kim2009} proposed using the check function as a possible extension of $\ell_1$-trend filtering but did not investigate it further. Here, we develop the basic $\ell_1$-quantile-trend-filtering framework and extend it to model multiple quantiles simultaneously with non-crossing constraints to ensure valid trend estimates. 
\item To reduce computation time and extend the method to long time series, we develop a parallelizable ADMM algorithm. 
\item Finally, we propose a modified criterion for performing model selection.
\end{itemize}

%% ----------------------------------------------------------------------
%% Methods
%% ----------------------------------------------------------------------
\section{Baseline Trend Estimation}
\label{sec:methods}

%We review \textit{$\ell_1$-trend filtering} for estimating a mean trend before discussing how to adapt it to estimate a baseline trend. We then describe a new scalable algorithm for simultaneously computing multiple quantile trends over large times windows.

%% ----------------------------------------------------------------------
%% Background
%% ----------------------------------------------------------------------
\subsection{Background}

\cite{Kim2009} originally proposed \textit{$\ell_1$-trend filtering} to estimate trends with piecewise polynomial functions, assuming that the observed time series $\V{y}$ consists of a trend $\V{\theta}$ plus uncorrelated noise $\V{\varepsilon}$, namely $\V{y} = \V{\theta} + \V{\varepsilon}$. The estimated trend is the solution to the following convex optimization problem
\begin{eqnarray*}
\underset{\V{\theta}}{\text{minimize}}\; \frac{1}{2} \lVert \V{y} - \V{\theta} \rVert_2^2 + \lambda \lVert \Mn{D}{k+1}\V{\theta} \rVert_1,
\end{eqnarray*}
	where $\lambda$ is a nonnegative regularization parameter, and the matrix $\Mn{D}{k+1} \in \Real^{(n - k -1) \times n}$ is the discrete difference operator of order $k+1$. To understand the purpose of penalizing the size of $\Mn{D}{k+1}\V{\theta}$ consider the difference operator when $k = 0$.

	\begin{eqnarray*}
	\Mn{D}{1} & = & \begin{pmatrix}
	-1 & 1 & 0 & \cdots & 0 & 0 \\
	0 & -1 & 1 & \cdots & 0 & 0 \\
	\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
	0 & 0 & 0 & \cdots & -1 & 1 \\
	\end{pmatrix}.
	\end{eqnarray*}

	Thus, $\lVert \Mn{D}{1}\V{\theta} \rVert_1 = \sum_{i=1}^{n-1} \lvert \theta_i - \theta_{i+1} \rvert$, which is known as total variation denoising penalty in one dimension in the signal processing literature \citep{Rudin1992} or the fused lasso penalty in the statistics literature \citep{Tibshirani2005}. The penalty term incentivizes solutions which are piecewise constant. For $k \geq 1$, the difference operator $\Mn{D}{k+1} \in \Real^{(n-k-1) \times n}$ is defined recursively as follows
	\begin{eqnarray*}
	\Mn{D}{k+1} & = & \Mn{D}{1}\Mn{D}{k}.
	\end{eqnarray*}
	Penalizing the 1-norm of the vector $\Mn{D}{k+1}\V{\theta}$ produces estimates of $\V{\theta}$ that are piecewise polynomials of order $k$.

@@

	\cite{Tib2014} proved that with a judicious choice of $\lambda$ the trend filtering estimate converges to the true underlying function at the minimax rate for functions whose $k$th derivative is of bounded variation and showed that trend filtering is both fast and locally adaptive when the time series consists of only the trend and random noise, which is illustrated in \Fig{tpn}. As noted earlier though in some applications, such as the air quality monitoring problem considered in this paper, the data contain a rapidly varying signal in addition to the slowly varying trend and noise. \Fig{tpspn} shows that standard trend filtering is not designed to distinguish between the slowly varying trend and the rapidly-varying signal, as the smooth component estimate $\V{\theta}$ is biased towards the peak of the transient component.

	To account for the presence of a transient component in the observed time series $\V{y}$, we propose quantile trend filtering \Fig{tpspn}. % as a more robust method for detrending in these situations \Fig{tpspn}. 
	To estimate the trend in the $\tau$\textsuperscript{th} quantile, we solve the convex optimization problem
	\begin{eqnarray}
	\label{eq:quantile_trend}
	\underset{\V{\theta}}{\text{minimize}}\; \rho_\tau(\V{y} - \V{\theta}) + \lambda \lVert \Mn{D}{k+1} \V{\theta} \rVert_1,
	\end{eqnarray}
	where $\rho_\tau(\V{r})$ is the check function
	\begin{eqnarray}
	\label{eq:check}
	\rho_{\tau}(\V{r}) & = & \sum_{i=1}^n \VE{r}{i}(1-\One(\VE{r}{i}<0)),
	\end{eqnarray}
	and $\One(A)$ is 1 if its input $A$ is true and 0 otherwise.

	\begin{figure}
		\centering
		\subfloat[Slow Trend + Noise]{\label{fig:tpn}
		\includegraphics[width = 0.45\linewidth]{Figures/trend_filter_eg1.png}}
		\subfloat[Slow Trend + Spikes + Noise]{\label{fig:tpspn}
		\includegraphics[width = 0.45\linewidth]{Figures/trend_filter_eg2.png}}
		\caption{Examples of trend filtering solutions (red) and 15\textsuperscript{th} quantile trending filtering solution (blue). Standard trend filtering performs well in the no-signal case (right) but struggles to distinguish between the slowly varying trend and the rapidly-varying signal (left). The quantile trend is not affected by the signal and provides an estimate of the baseline.}
	\end{figure}
	Note that we do not explicitly model $\V{s}$. Rather, we focus on estimating $\V{\theta}$. We then estimate $\V{s} + \V{\varepsilon}$ as the difference $\V{y} - \V{\theta}$.

	Before elaborating on how we compute our proposed $\ell_1$-quantile trend filtering estimator, we discuss similarities and differences between our proposed estimator and existing baseline estimators.

%% ----------------------------------------------------------------------
%% Relationship to Prior Work
%% ----------------------------------------------------------------------
\subsubsection{Relationship to Prior Work}

@@

	\begin{equation*}
	\label{eq:smoothingspline}
	\sum_{i=1}^n\rho_{\tau}(y(t_i) - \theta(t_i)) + \lambda\int \theta''(t)^2 dt, 
 	\end{equation*}	
	where $\theta(t)$ is a smooth function of time and $\lambda$ is a tuning parameter that controls the degree of smoothing. \cite{Oh2011} proposed an algorithm for solving the quantile smoothing spline problem by approximating the check function with a differentiable function. \cite{Racine2017} take a different approach and constrain the response to follow a smooth location scale model of the form $y(t) = a(t) + b(t)\epsilon_i$ and estimate the $\tau$\textsuperscript{th} conditional quantile using a kernel estimator and local linear approach. \Halley{Eric}{What are $a(t), b(t)$, and $\epsilon_i$?}

%% ----------------------------------------------------------------------
%% Quantile Trend Filtering
%% ----------------------------------------------------------------------
\subsection{Quantile Trend Filtering}

	We combine the ideas of quantile regression and trend filtering. For a single quantile level $\tau$, the quantile trend filtering problem is given in \Eqn{quantile_trend}. As with classic quantile regression, the quantile trend filtering problem is a linear program which can be solved by a number of methods. In many cases, including ours, we are interested in estimating multiple quantiles simultaneously. We also want to ensure that our quantile estimates are valid by enforcing the constraint that if $\tau_2 > \tau_1$ then $Q(\tau_2) \ge Q(\tau_1)$ where $Q$ is the quantile function of $\V{y}$. Given quantiles $\tau_1 < \tau_2 < \ldots < \tau_J$, the optimization problem becomes 
	\begin{eqnarray*}
	\label{eq:noncross_trend}
%	&&\underset{\V{\theta}_1, \ldots, \V{\theta}_J}{\text{minimize}}\; \sum_{j=1}^J \left [\rho_{\tau_j}(\V{y} - \V{\theta}_{j}) + 
%	\lambda_j \lVert \Mn{D}{k+1} \V{\theta}_j \rVert_1 \right ] \\	
	&&\underset{\M{\Theta} \in \mathcal{C}}{\text{minimize}}\; \sum_{j=1}^J \left [\rho_{\tau_j}(\V{y} - \V{\theta}_{j}) + 
	\lambda \lVert \Mn{D}{k+1} \V{\theta}_j \rVert_1 \right ] \\
%	&&\text{subject to: }\; \V{\theta}_1 \leq \V{\theta}_2 \leq \ldots \leq \V{\theta}_J	
%	\theta_{1}(t) \le \theta_{2}(t) \le \ldots \le \theta_{J}(t) \text{ for all } t, \nonumber
	\end{eqnarray*}
	where $\M{\Theta} = \begin{pmatrix} \V{\theta}_1 & \V{\theta}_2 & \cdots & \V{\theta}_J \end{pmatrix} \in \Real^{n \times J}$ is a matrix whose $j$th column corresponds to the $j$th quantile signal $\V{\theta}_j$ and the set $\mathcal{C} = \{\M{\Theta} \in \Real^{n \times J}: \ME{\theta}{ij} \leq \ME{\theta}{ij'} \;\text{for $j \leq j'$}\}$ encodes the non-crossing quantile constraints. The additional constraints are linear in the parameters, so the non-crossing quantile trends can still be estimated by a number of available solvers. In the rest of this paper, we report numerical results using the commercial solver Gurobi \citep{gurobi} and its R package implementation. However, we could easily substitute a free solver such as the Rglpk package by \cite{rglpk}.

@@

	The number of parameters to be estimated in this problem is equal to the number of observations multiplied by the number of quantiles of interest. As the size of the data and the number of quantiles grows, all solvers will eventually break \Halley{Eric}{Can we be more specific about what we mean when we say ``break?"}. To our knowledge, no one has addressed the problem of finding smooth quantile trends of series that are too large to be processed simultaneously. We propose a divide-and-conquer approach via an ADMM algorithm for solving large problems in a piecewise fashion. The ADMM algorithm \citep{gabay1975dual, glowinski1975approximation} is  described in greater detail by \cite{boyd2011distributed}, but we briefly review how it can be used to iteratively solve the following equality constrained optimization problem.
\begin{equation}
\label{eq:split_objective}
\begin{split}
&\text{minimize} \; f(\V{\theta}) + g(\Vtilde{\theta}) \\
&\text{subject to} \; \M{A}\V{\theta} + \M{B}\Vtilde{\theta} = \V{c}.
\end{split}
\end{equation}
Recall that finding the minimizer to an equality constrained optimization problem is equivalent to the identifying the saddle point of the Lagrangian function associated with the problem \Eqn{split_objective}. ADMM seeks the saddle point of a related function called the augmented Lagrangian,
%The Lagrangian for the ALM problem
\begin{eqnarray*}
\mathcal{L}_{\gamma}(\V{\theta},\Vtilde{\theta},\V{\omega}) & = & f(\V{\theta}) + g(\Vtilde{\theta}) + \langle \V{\omega}, \V{c} - \M{A}\V{\theta} - \M{B}\Vtilde{\theta} \rangle
+ \frac{\gamma}{2} \lVert \V{c} - \M{A}\V{\theta} - \M{B}\Vtilde{\theta} \rVert_2^2,
\end{eqnarray*}
where the dual variable $\V{\omega}$ is a vector of Lagrange multipliers and $\gamma$ is a nonnegative tuning parameter. When $\gamma = 0$, the augmented Lagrangian coincides with the ordinary Lagrangian.

ADMM minimizes the augmented Lagrangian one block of variables at a time before updating the dual variable $\V{\omega}$. This yields the following sequence of updates at the $m+1$th ADMM iteration
\begin{equation}
\label{eq:admm_updates}
\begin{split}
\V{\theta}_{m+1} & \amp = \amp \underset{\V{\theta}}{\arg\min}\; \mathcal{L}_\gamma(\V{\theta},\Vtilde{\theta}_{m}, \V{\omega}_{m}) \\
\Vtilde{\theta}_{m+1} & \amp = \amp \underset{\Vtilde{\theta}}{\arg\min}\; \mathcal{L}_\gamma(\V{\theta}_{m+1},\Vtilde{\theta}, \V{\omega}_{m}) \\
\V{\omega}_{m+1} & \amp = \amp \V{\omega}_{m} + \gamma(\V{c} - \M{A}\V{\theta}_{m+1} - \M{B}\Vtilde{\theta}_{m+1}).
\end{split}
\end{equation}

	Let $\mathcal{I}$ denote a subset of the indices $\{1, \ldots, n\}$ and $\V{y}_\mathcal{I} \in \Real^{\lvert I \rvert}$ denote the subvector obtained by stacking on top of each other the elements $\VE{y}{i}$ for $i \in \mathcal{I}$ in their canonical ordering.	
	We apply the consensus ADMM algorithm to the quantile regression trend filtering problem given in \Eqn{quantile_trend}, by dividing our observed series $\VE{y}{i}$ with $i = 1, \ldots, n$ into overlapping $W$ windows of observations $\Vn{y}{w} = \V{y}_{\mathcal{I}_w}$ where $\mathcal{I}_w$ is the set of integers from $l_w$ to $u_w$ inclusive, namely $\mathcal{I}_w = \{l_w, l_w + 1, \ldots, u_w -1, u_w\}$ where 
\begin{eqnarray*}
1 = l_{1} < l_{2} < u_{1} < l_{3} < u_{2} < l_{4} < u_{3} < \cdots < u_{W} = n.
\end{eqnarray*}
\Fig{windows} shows an example of 1200 observations being mapped into three equally sized overlapping windows of observations.
